---
title: "Univariate Analysis"
author: "João Brandão"
date: "2023-04-24"
output: html_document
---

# Load Packages

```{r load_packages, message = FALSE}

library(survival)
library(qvalue)

```


# Log-Rank Test

Excluding genes (variables) based on the log-rank test for univariable models with each gene.
An alpha of 0.20 was considered.

Firstly, each observation is going to be divided into two groups:

1st group) gene expression > median(gene expression) --> High Risk
2nd group) gene expression < median(gene expression) --> Low Risk

## Prepare Data

```{r prapare_data}

survival_df <- ydata
survival_df$status <- as.numeric(ydata$status) - 1

group_median <- matrix(nrow = dim(xdata)[1], ncol = dim(xdata)[2])
colnames(group_median)<-colnames(xdata[,1:dim(xdata)[2]])
  
```

## Obtaining the low and the high risk for each concentration

```{r risk matrix}

for (i in 1:dim(xdata)[2]){
 group_median[,i] <- ifelse(xdata[,i] > median(xdata[,i]), "high", "low")
}


# transform into a data frame

group_median_df<-as.data.frame(group_median)

for (i in 1:dim(group_median_df)[2]){
  group_median_df[,i] <- as.factor(group_median_df[,i])
} 

head(group_median_df)

```

## Performing the log-rank test

```{r log_rank_test}

var.to.test <- names(group_median_df)   #variables of interest
nvar <- length(var.to.test)
chisq <- double(nvar)
bad.genes <- c()

for (i in 1:dim(xdata)[2]) {
     tfit <- tryCatch(survdiff(Surv(survival_df$time,survival_df$status) ~ get(var.to.test[i]), data=group_median_df),
                      error = function(msg) { 
                        bad.gene <- colnames(xdata)[i]
                        #flog.info('problem whith i = %d called %s', i, bad.gene)
                        bad.genes <<- c(bad.genes, bad.gene)
                        })
     chisq[i] <- tfit$chisq
     }
  
```

# Remove Bad Genes 

```{r remove_genes}

pvalue <- double(nvar)
for (i in 1:dim(xdata)[2]){
    pvalue[i] = p.adjust(    
      pchisq(chisq[i], df=1, lower.tail=FALSE), # adjusting the p-values 
      method = "BH" # Benjamini & Hochberg fdr
    )
}

pvalue <- as.numeric(round(pvalue, digits = 5))
length(which(pvalue<0.20)) # BH - 5315; none - 5315


# New covariates dataset

removed.genes <- as.vector(which(pvalue > 0.20))

xdata.reduced <- xdata[, - removed.genes]
dim(xdata.reduced)
head(xdata.reduced)


data.reduced <- data[, - (removed.genes + 2)]
head(data.reduced)

```

```{r}

data <- data[, - (removed.genes + 2)] 
head(data)

```
