---
title: "Variable Selection - Elastic Net"
author: "João Brandão"
date: "2024-04-25"
output: html_document
---

# Loading Packages

```{r load_packages, message = FALSE}

library(survival)
library(glmSparseNet)
library(glmnet)

```


# Variable Selection

```{r elastic_net}

start.time <- Sys.time()

my.alpha <- seq(0.1,0.9,0.1)
lambda.1se.EN    <- c()
cv_coef.EN       <- c()
cv_var.EN        <- c()
cv_genes.EN      <- c()
c_index_train.EN <- c()
c_index_test.EN  <- c()
cv_results.EN    <- c()
fit.cv.EN <- c()

for (i in 1:length(my.alpha)){
  set.seed(11)
  # cross validation
  fit.cv.EN[i] <- list(cv.glmnet(xdata.train, ydata.train, 
                         family  = 'cox', 
                         type.measure = "C",
                         nlambda = 1000,
                         alpha = my.alpha[i],
                         nfolds = 10
                         ))
  plot(fit.cv.EN[[i]])
  
  lambda.1se.EN[i] <- fit.cv.EN[[i]]$lambda.min
  cv_coef.EN[i]    <- list(coef(fit.cv.EN[[i]])[which(coef(fit.cv.EN[[i]], s='lambda.min') != 0)])
  cv_var.EN[i]     <- list(xdata.train[, which(coef(fit.cv.EN[[i]], s='lambda.min') != 0)])
  cv_genes.EN[i]   <- list(colnames(xdata.train[, which(coef(fit.cv.EN[[i]], s='lambda.min') != 0)]))
  
  c_index_train.EN[i] <- fit.cv.EN[[i]]$cvm[fit.cv.EN[[i]]$lambda == fit.cv.EN[[i]]$lambda.min]
  pred <- predict(object = fit.cv.EN[[i]], xdata.test, type="response")
  c_index_test.EN[i] <- apply(pred, 2, Cindex, y = ydata.test)
  cv_results.EN[i] <- list(pred)
}

end.time <- Sys.time()
time.taken <- end.time - start.time
time.taken

```


# Survival curves

```{r survival_curves}

sc_train.EN <- list()
sc_test.EN  <- list()


for (i in 1:length(my.alpha)){
title_train <- paste0("Survival Curve Train Data - EN: Alpha = ", my.alpha[i])  
title_test  <- paste0("Survival Curve Test Data - EN: Alpha = ", my.alpha[i])

# Train
sc_train.EN[[i]] <- separate2GroupsCox(as.vector(cv_coef.EN[i]), 
                                       xdata.train[, unlist(cv_genes.EN[i], use.names=FALSE)], 
                                       y_train, 
                                       plot.title = title_train, 
                                       legend.outside = FALSE)

# Test
sc_test.EN[[i]] <- separate2GroupsCox(as.vector(cv_coef.EN[i]),
                                    xdata.test[, unlist(cv_genes.EN[i], use.names=FALSE)], 
                                    y_test, 
                                    plot.title = title_test, 
                                    legend.outside = FALSE)
}

sc_train.EN ;sc_test.EN


```


# Log-Rank Test

```{r log_rank_test}

pv_train <- c()
pv_test  <- c()

for (i in 1:length(my.alpha)){
  pv_train[i] <- sc_train.EN[[i]]$pvalue
  pv_test[i] <- sc_test.EN[[i]]$pvalue
}
log_rank.EN <- cbind(my.alpha, pv_train, pv_test)

log_rank.EN <- data.frame(log_rank.EN)
colnames(log_rank.EN) <- c("Alpha Used", "P-Value Train Set", "P-Value Test Set")
log_rank.EN

```


## Convert to coxph

With the cross validation results the models are converted into coxph objects. This is a necessary step for computing the proportional hazards test.

```{r coxph}

fit_EN <- c() # Cox models 
y <- Surv(y_train$time, as.numeric(y_train$status) - 1) # response variable for coxph

for (i in 1:length(my.alpha)){
  tryCatch({
  fit_EN[i] <- list(
    coxph(y ~ ., data = xdata.train[, cv_genes.EN[[i]]],
                init = as.numeric(cv_coef.EN[[i]]), # specify coefficient values
                iter.max = 0 # force the software to keep those values
  ))
  }, error=function(e){})
}

model.name_EN <- c()
for (i in 1:length(my.alpha)) model.name_EN[i] <- paste0("EN Model ", i)  

table <- cbind(model.name_EN, my.alpha, lengths(cv_genes.EN))
colnames(table) <- c("Model", "Alpha", "Covariates")
data.frame(table)

```


## Concordance Index (C-index)

The C-index represents the model’s ability to correctly provide a reliable ranking of the survival times based on the individual risk scores. Similarly to the AUC, a value of 1 corresponds to the best model prediction, and a value of 0.5 represents a random prediction. 
[Reference:
https://square.github.io/pysurvival/metrics/c_index.html#:~:text=The%20concordance%20index%20or%20C-index%20is%20a%20generalization,survival%20times%20based%20on%20the%20individual%20risk%20scores.]

```{r c_index}

c.index.train_EN <- c() # c-index of the train data set 
c.index.test_EN  <- c() # c-index of the test data set

for (i in 1:length(my.alpha)){
  c.index.train_EN[i] <- fit.cv.EN[[i]]$cvm[fit.cv.EN[[i]]$lambda == fit.cv.EN[[i]]$lambda.min]
  pred <- predict(object = fit.cv.EN[[i]], xdata.test, type="response")
  c.index.test_EN[i] <- apply(pred, 2, Cindex, y = ydata.test) 
}

table <- cbind(round(c.index.train_EN, 4), round(c.index.test_EN, 4))
colnames(table) <- c("C.Index_train", "C.Index_test")
data.frame(table)

```

# Overall Results

```{r result_table}

table_EN <- cbind(model.name_EN, my.alpha, round(lambda.1se.EN, 4), lengths(cv_genes.EN), 
                  round(c.index.train_EN, 4), round(c.index.test_EN, 4), 
                  round(pv_train, 4), round(pv_test, 4))
colnames(table_EN) <- c("Model", "Alpha", "Lambda", "Covariates",
                        "C.index_train", "C.index_test",
                        "LR.pvalue_train", "LR.pvalue_test")
(table_EN <- data.frame(table_EN))

```

Best selection: 6 
