---
title: "Martingale Residuals"
author: "João Brandão"
date: "2024-03-25"
output: html_document
---

# Load Packages

```{r}

library(ggplot2)
library(ggrepel)
library(dplyr)
library(survival)

```


# Elastic Net

## Define Cox models

The cox model is defined for all observations (training + test) so that the residuals for all observations are obtained.


```{r cox_model}

y <- Surv(ydata$time, as.numeric(ydata$status) - 1) # response variable for coxph

coxph.EN <- coxph(y ~ ., data = xdata[, genes.EN],
                init = as.numeric(cox.EN$coefficients), # specify coefficient values
                iter.max = 0 # force the software to keep those values
                )

```


## Calculate Martingale Residuals 

```{r}

# cox.EN$residuals == res.EN
res.EN <- residuals(cox.EN, type= "martingale")

```

## Plot Residuals

### Theme

```{r theme_}

My_Theme <- theme(
  plot.title = element_text(size = 11, face="bold"),
  axis.title.x = element_text(size = 10),
  axis.text.x = element_text(size = 10),
  axis.title.y = element_text(size = 10),
  axis.text.y = element_text(size = 10))

```

### Scatter Plot

```{r plot_residuals}

n <- 10 # set to find the top n residuals

title <- "Plot of residuals | Elastic Net - Model 6"

# Data frame with the residual values of the respective observations
df <- data.frame(cbind(1:length(res.EN), rownames(ydata), as.vector(res.EN)))
colnames(df) <- c("index", "names", "residuals")
df$index <- as.numeric(df$index)
df$residuals <- as.numeric(df$residuals)

# Indices of the observations with the highest residuals (absolute value)
highest.indices <- order(abs(res.EN), decreasing = TRUE)[1:n]


res_plot.EN <- ggplot(df, aes(x = index, y = residuals)) + 
  geom_point() + 
  geom_text_repel(aes(label = ifelse(index %in% highest.indices, names, '')),  segment.color = 'grey50', hjust = - 0.5, size = 3) +
  theme_classic() + ggtitle(title) + ylab("Residuals") + My_Theme
  
res_plot.EN

```

### Box Plot

```{r boxplot_residuals}

title <- "BoxPlot of residuals | Elastic Net - Model 6"

# Function to find the points outside the box plot whiskers
find_outlier <- function(x) {
  return(x < quantile(x, .25) - 1.5*IQR(x) | x > quantile(x, .75) + 1.5*IQR(x))
}

# Data frame with the residual values of the respective observations
df <- data.frame(cbind(1:length(res.EN[[i]]), rownames(ydata), as.vector(res.EN)))
colnames(df) <- c("index", "names", "residuals")
df$index <- as.numeric(df$index)
df$residuals <- as.numeric(df$residuals)


# Add new column to data frame that indicates if each observation is an outlier
df <- df %>%
        mutate(outlier = ifelse(find_outlier(residuals), names, NA))


res_boxplot.EN <- ggplot(df, aes(x = " ", y = residuals)) +
  geom_boxplot() +
  geom_text_repel(aes(label = outlier), na.rm=TRUE, hjust=-.0, size = 2.5) +
  theme_classic() + ggtitle(title) + ylab("Residuals") + xlab("") + My_Theme + theme(aspect.ratio = 1)

res_boxplot.EN

```


# Hub Cox

## Define Cox models

The cox model is defined for all observations (training + test) so that the residuals for all observations are obtained.


```{r}

y <- Surv(ydata$time, as.numeric(ydata$status) - 1) # response variable for coxph

coxph.HC <- coxph(y ~ ., data = xdata[, genes.HC],
                init = as.numeric(cox.HC$coefficients), # specify coefficient values
                iter.max = 0 # force the software to keep those values
                )

```


## Calculate Martingale Residuals 

```{r}

res.HC <- residuals(cox.HC, type= "martingale")
summary(res.HC)

```

## Plot Residuals

### Theme

```{r theme_}

My_Theme <- theme(
  plot.title = element_text(size = 11, face="bold"),
  axis.title.x = element_text(size = 10),
  axis.text.x = element_text(size = 10),
  axis.title.y = element_text(size = 10),
  axis.text.y = element_text(size = 10))

```

### Scatter Plot

```{r plot_residuals}

n <- 10 # set to find the top n residuals

title <- "Plot of residuals | Hub Cox - Model 9"

# Data frame with the residual values of the respective observations
df <- data.frame(cbind(1:length(res.HC), rownames(ydata), as.vector(res.HC)))
colnames(df) <- c("index", "names", "residuals")
df$index <- as.numeric(df$index)
df$residuals <- as.numeric(df$residuals)

# Indices of the observations with the highest residuals (absolute value)
highest.indices <- order(abs(res.HC), decreasing = TRUE)[1:n]


res_plot.HC <- ggplot(df, aes(x = index, y = residuals)) + 
  geom_point() + 
  geom_text_repel(aes(label = ifelse(index %in% highest.indices, names, '')),  segment.color = 'grey50', hjust = - 0.5, size = 3) +
  theme_classic() + ggtitle(title) + ylab("Residuals") + My_Theme
  
res_plot.HC

```

### Box Plot

```{r boxplot_residuals}

title <- "BoxPlot of residuals | Hub Cox - Model 9"

# Function to find the points outside the box plot whiskers
find_outlier <- function(x) {
  return(x < quantile(x, .25) - 1.5*IQR(x) | x > quantile(x, .75) + 1.5*IQR(x))
}

# Data frame with the residual values of the respective observations
df <- data.frame(cbind(1:length(res.HC[[i]]), rownames(ydata), as.vector(res.HC)))
colnames(df) <- c("index", "names", "residuals")
df$index <- as.numeric(df$index)
df$residuals <- as.numeric(df$residuals)


# Add new column to data frame that indicates if each observation is an outlier
df <- df %>%
        mutate(outlier = ifelse(find_outlier(residuals), names, NA))


res_boxplot.HC <- ggplot(df, aes(x = " ", y = residuals)) +
  geom_boxplot() +
  geom_text_repel(aes(label = outlier), na.rm=TRUE, hjust=-.0, size = 2.5) +
  theme_classic() + ggtitle(title) + ylab("Residuals") + xlab("") + My_Theme + theme(aspect.ratio = 1)

res_boxplot.HC

```

# Orphan Cox

## Define Cox models

The cox model is defined for all observations (training + test) so that the residuals for all observations are obtained.


```{r}

y <- Surv(ydata$time, as.numeric(ydata$status) - 1) # response variable for coxph

coxph.OC <- coxph(y ~ ., data = xdata[, genes.OC],
                init = as.numeric(cox.OC$coefficients), # specify coefficient values
                iter.max = 0 # force the software to keep those values
                )

```


## Calculate Martingale Residuals 

```{r}

res.OC <- residuals(cox.OC, type= "martingale")
summary(res.OC)

```

## Plot Residuals

### Theme

```{r theme_}

My_Theme <- theme(
  plot.title = element_text(size = 11, face="bold"),
  axis.title.x = element_text(size = 10),
  axis.text.x = element_text(size = 10),
  axis.title.y = element_text(size = 10),
  axis.text.y = element_text(size = 10))

```

### Scatter Plot

```{r plot_residuals}

n <- 10 # set to find the top n residuals

title <- "Plot of residuals | Orphan Cox - Model 7"

# Data frame with the residual values of the respective observations
df <- data.frame(cbind(1:length(res.OC), rownames(ydata), as.vector(res.OC)))
colnames(df) <- c("index", "names", "residuals")
df$index <- as.numeric(df$index)
df$residuals <- as.numeric(df$residuals)

# Indices of the observations with the highest residuals (absolute value)
highest.indices <- order(abs(res.OC), decreasing = TRUE)[1:n]


res_plot.OC <- ggplot(df, aes(x = index, y = residuals)) + 
  geom_point() + 
  geom_text_repel(aes(label = ifelse(index %in% highest.indices, names, '')),  segment.color = 'grey50', hjust = - 0.5, size = 3) +
  theme_classic() + ggtitle(title) + ylab("Residuals") + My_Theme
  
res_plot.OC

```

### Box Plot

```{r boxplot_residuals}

title <- "BoxPlot of residuals | Orphan Cox - Model 7"

# Function to find the points outside the box plot whiskers
find_outlier <- function(x) {
  return(x < quantile(x, .25) - 1.5*IQR(x) | x > quantile(x, .75) + 1.5*IQR(x))
}

# Data frame with the residual values of the respective observations
df <- data.frame(cbind(1:length(res.OC[[i]]), rownames(ydata), as.vector(res.OC)))
colnames(df) <- c("index", "names", "residuals")
df$index <- as.numeric(df$index)
df$residuals <- as.numeric(df$residuals)


# Add new column to data frame that indicates if each observation is an outlier
df <- df %>%
        mutate(outlier = ifelse(find_outlier(residuals), names, NA))


res_boxplot.OC <- ggplot(df, aes(x = " ", y = residuals)) +
  geom_boxplot() +
  geom_text_repel(aes(label = outlier), na.rm=TRUE, hjust=-.0, size = 2.5) +
  theme_classic() + ggtitle(title) + ylab("Residuals") + xlab("") + My_Theme + theme(aspect.ratio = 1)

res_boxplot.OC

```
