---
title: "RP test"
author: "João Brandão"
date: "2024-03-25"
output: html_document
---


# Load packages

```{r packages, message = FALSE}

library(qvalue)
library(ggplot2)
library(ggrepel)

```


# Rank Products Test

In order to obtain a global result concerning each one of the techniques performed in the previous section, a rank product test is going to be used. The RP Test is performed between the best models of each regularization method. The criterion for best model is firstly the C-index and then the p-value for the log-rank test (both corresponding to the test data set). First a rank matrix is going to be obtained based on the ranks of each outliers detection method.


## Rank Matrixes

```{r rank_matrix}

rank.martingale_EN <- rank(-abs(as.vector(res.EN)), ties.method = "first")
rank.martingale_HC <- rank(-abs(as.vector(res.HC)), ties.method = "first")
rank_martingale.OC <- rank(-abs(as.vector(res.OC)), ties.method = "first")

rankMat <- cbind(rank.martingale_EN, rank.martingale_HC, rank_martingale.OC)
               
```


## Rank Product

```{r rank_product}

RankProduct = function(rankMat) {
  return(apply(rankMat, 1, prod))
}

rankproduct <- RankProduct(rankMat)

```


## P-Values

In order to obtain the p-values the algorithm proposed by Heskes (2014) was used. Notice that the p-values are based on the geometric mean of upper and lower bounds, defined recursively.

The input to obtain the p-values is the following:
+ rho: the rank product matrix;
+ the number of observations, n;
+ the number of methods used, k.

```{r pvalues}

# input
rho = rankproduct
n <- dim(xdata)[1]
k <- dim(rankMat)[2]

pvaluesRP <- as.vector(rankprodbounds(rho, n, k, Delta ='geometric'))

# obtaining the observations with the lowest p-values
pvaluesRP.matrix <- cbind(pvaluesRP)
which(pvaluesRP.matrix < 0.01) # pvalues < to 1%
which(pvaluesRP.matrix < 0.05) # pvalues < to 5%

```


## Q-Values

Perform an adjustment of the rank products p-value. The q-value of a test measures the proportion of false positives incurred (called the false discovery rate) when that particular test is called significant.

```{r qvalues}

qvaluesRP <- p.adjust(pvaluesRP, method = "BH")

which(qvaluesRP < 0.01)
which(qvaluesRP < 0.05)

```


# Top Outliers

```{r top_outliers}

outliers <- as.data.frame(cbind(ydata$time,(as.numeric(ydata$status) - 1), res.EN,
                                     rank.martingale_EN, 
                                     rank.martingale_HC, 
                                     rank_martingale.OC,
                                     round(pvaluesRP, 4), round(qvaluesRP,4)))

colnames(outliers) <- c("Survival Time", "Status", "Rank EN", "Rank HC", "Rank OC", "P-Value", "Q-Value") 
rownames(outliers) <- rownames(ydata)

# Top outliers
sort.outliers <- outliers[order(qvaluesRP) ,]
sort.outliers

```

# Distribution Plot

```{r outlier_distribution}

q <- 0.05 ## Q-value considered ##

title <- ""

df <- ydata[order(ydata$time),]
df <- cbind(df, rownames(df), 1:length(rownames(df)))
df$status <- as.numeric(df$status) - 1
colnames(df) <- c("Time", "Status", "Names", "Index")

out <- rownames(outliers)[which(qvaluesRP < q)]
highlight1_df <- df[which(df$Names %in% out),]
highlight1_df <- highlight1_df[which(highlight1_df$Status == 1),]
highlight0_df <- df[which(df$Names %in% out),]
highlight0_df <- highlight0_df[which(highlight0_df$Status == 0),]


out_plot <- ggplot(df, aes(x = Index, y = Time)) + 
  geom_point(alpha = 0.2, size = 1.5) + 
  geom_text_repel(aes(label = ifelse(Names %in% rownames(outliers)[which(qvaluesRP < q)], Names, '')),  segment.color = 'grey50', hjust = - 0.7, size = 3) +
  geom_point(data = highlight1_df, 
             aes(x = Index, y = Time, color="Died"), size = 2) +
  geom_point(data = highlight0_df, 
             aes(x = Index, y = Time, color="Lost to follow-up"), size = 2) +
  scale_color_manual(name="Patient Status", 
                     breaks=c("Died", "Lost to follow-up"),
                     values=c("Died"='#ff2800', "Lost to follow-up"='#0000cd')) +
  theme_bw() + ggtitle(title) + ylab("Survival Time (Days)") + xlab("Patient Index") + My_Theme +
  theme(legend.position = c(.215, .79), legend.title = element_text(size=8), legend.text = element_text(size=8)) +
  scale_x_continuous(expand = c(0, 0), limits = c(0, 210)) + 
  scale_y_continuous(expand = c(0, 0), limits = c(0, 2500))
  
out_plot
 
```
